---
title: "Bayesian Kernel Machine Regression (BKMR)"
format: html
---

## What question does this answer?
Is there a relationship between the overall mixture and my (health) outcome? 
Which components in the mixture are driving the outcome? 
Same as with WQSR, but without the assumptions.


## When can I use it?
When you have mixture components that may have complex non-linear and non-additive
relationships with health


## How do I use it?

- Enable variable selection  
- If you are aware of highly collinear elements in the mixture, add this knowledge
to the model (by specifying which groups of elements are collinear)  
- Tune the model!  


## Tuning

TBA  
https://jenfb.github.io/bkmr/overview.html  
https://cran.r-project.org/web/packages/bkmrhat/vignettes/bkmrhat-vignette.html  
The Metropolis–Hastings tuning parameters were adjusted to achieve an acceptance rate of 20–50%

## Reporting
Now that you have tuned your model, report your findings. For now, these
assume a continuous outcome

::: {.panel-tabset}

## Without collinearity
### Without highly collinear elements (no groups):

```{r}
#| label: bkmr_no_groups_example

library(ggplot2)
set.seed(111)
dat <- bkmr::SimData(n = 50, M = 4)
y <- dat$y
Z <- dat$Z
X <- dat$X

fitkm_comp <- bkmr::kmbayes(y = y, # vector of outcomes
                       Z = Z, # matrix of exposures, each col being an exposure
                       X = X, # matrix of covariates, each col is a covariate
                       iter = 10000, # should be 30k-60k 
                       family = "gaussian",
                       verbose = FALSE, 
                       varsel = TRUE, # variable selection enabled
                       # without groups it's component-wise variable selection 
                       # if I wanted to use fine-tuned parameters e.g:
                       # control.params = list(r.jump1 = 0.5, r.jump2 = 0.5)
                       # r.jump if varsel=FALSE 
                       )

# The posterior inclusion probabilities (PIPs) cleary indicate that z1 and z2 
# should be included in the model (strong evidence would be >= 90 or 95 even)

bkmr::ExtractPIPs(fitkm_comp)


# We can explore the univariate relationship between each exposure and the outcome,
# where all of the other exposures are fixed to a particular percentile and all 
# covariates are held constant. The argument specifying the quantile at which to
# fix the other exposures is given by q.fixed. 
# Here we see that for z1 and z2, increasing values of the exposures is associated
# with increasing values of the outcome Y (for each of the other predictors in Z
# fixed to their 50th percentile). It also looks like z1 may have a nonlinear 
# relationship with the outcome

pred.resp.univar.comp <- bkmr::PredictorResponseUnivar(fit = fitkm_comp,
                                                        q.fixed = 0.5)

ggplot(pred.resp.univar.comp, 
       aes(z, est, ymin = est - 1.96*se, ymax = est + 1.96*se)
       ) + 
      geom_smooth(stat = "identity") + 
      facet_wrap(~ variable) +
      ylab("h(z)")


# We can explore the predictor-response function of a single predictor in Z for 
# the second predictor in Z fixed at various quantiles (and for the remaining 
# predictors fixed to a particular percentile). 
# The argument qs specifies a sequence of quantiles at which to fix the second 
# predictor.

pred.resp.bivar <- bkmr::PredictorResponseBivar(fit = fitkm_comp, min.plot.dist = 1)
pred.resp.bivar.levels <- bkmr::PredictorResponseBivarLevels(
                          pred.resp.df = pred.resp.bivar, 
                          Z = Z, 
                          qs = c(0.1, 0.5, 0.9))
# If the lines are parallel, there is no interaction between these two elements,
# if not, there is an interaction
ggplot(pred.resp.bivar.levels, aes(z1, est)) + 
    geom_smooth(aes(col = quantile), stat = "identity") + 
    facet_grid(variable2 ~ variable1) +
    ggtitle("h(expos1 | quantiles of expos2)") +
    xlab("expos1")


# We can compute the overall effect of the predictors, by comparing the value of 
# h (kernel function) when all of the predictors are at a particular percentile 
# and when all of them are at their 50th percentile. 
# The qs arg takes a sequence of values of quantiles and q.fixed the fixed quantile 
# (default is 50th percentile)

risks.overall <- bkmr::OverallRiskSummaries(fit = fitkm_comp, y = y, Z = Z, X = X, 
                                      qs = seq(0.25, 0.75, by = 0.05), 
                                      q.fixed = 0.5, method = "exact")
risks.overall
ggplot(risks.overall, aes(quantile, est, ymin = est - 1.96*sd, ymax = est + 1.96*sd)) + 
    geom_pointrange()


# Another summary of h that may be of interest would be to summarize the 
# contribution of an individual predictor to the response (single-predictor 
# health risks). For example, we may wish to compare risk when a single predictor
# in h is at the 75th percentile as compared to when that predictor is at its 
# 25th percentile, where we fix all of the remaining predictors to a particular 
# percentile. The two different quantiles at which to compare the risk are 
# specified using the qs.diff argument, and a sequence of values at which to
# fix the remaining pollutants can be specified using the q.fixed argument.
# The plot makes it easier to see.

risks.singvar <- bkmr::SingVarRiskSummaries(fit = fitkm_comp, y = y, Z = Z, X = X, 
                                      qs.diff = c(0.25, 0.75), #h(z_{q2}) - h(z_{q1})
                                      q.fixed = c(0.25, 0.50, 0.75),
                                      method = "exact")

# Here we see that a change in z1 from it's 25th to 75th percentile, where the 
# rest of the predictors are fixed at their 75th percentile (and the covariates x
# are constant), results in a ~1.68 estimated (additive) change in h(). 

# We also see the predictors z3 and z4 not contributing to the "risk" and that
# higher values of z1 and z2 are associated with higher values of h().
# The plot also suggests that for z1, as the remaining predictors increase in 
# value from their 25th to their 75th percentile, the risk of the outcome 
# associated with z1 increases. A similar pattern occurs for z2, which indicates
# the potential for interaction of z1 and z2
ggplot(risks.singvar, aes(variable, est, ymin = est - 1.96*sd, 
                          ymax = est + 1.96*sd, col = q.fixed)) + 
       geom_pointrange(position = position_dodge(width = 0.75)) + 
        coord_flip() +
        labs(x = "variable (from 25th to 75th percentile)")
  
 

# To make this notion a bit more formal, we may wish to compute specific ‘interaction’ 
# parameters. For example, we may want to compare the single-predictor health risks
# when all of the other predictors in Z are fixed to their 75th percentile to when
# all of the other predictors in Z are fixed to their 25th percentile. In the previous
# plot, this corresponds to subtracting the estimate represented by the red circle
# from the estimate represented by the blue circle. 

risks.int <- bkmr::SingVarIntSummaries(fit = fitkm_comp, y = y, Z = Z, X = X, 
                                 qs.diff = c(0.25, 0.75), # in z1
                                 qs.fixed = c(0.25, 0.75), # the rest of z
                                 method = "exact")
risks.int
 
# We see that single pollutant risk associated with a change in z1 from its 25th
# to 75th percentile increases by 0.7 when all other z are fixed at their 25th 
# percentile, as compared to when all other z are fixed at their 75th percentile.
 

```
## With collinearity
### With highly collinear elements


```{r}
#| label: bkmr_groups_example


set.seed(111)
d2 <- bkmr::SimData(n = 100, M = 4, Zgen = "corr", sigsq.true = 2.2)
y <- d2$y
Z <- d2$Z
X <- d2$X

fitkm_hier <- bkmr::kmbayes(y = y, # vector of outcomes
                       Z = Z, # matrix of exposures, each col being an exposure
                       X = X, # matrix of covariates, each col is a covariate
                       iter = 10000, # should be 30k-60k 
                       verbose = FALSE, 
                       family = "gaussian",
                       varsel = TRUE, # variable selection enabled
                       # with groups, it's hierarchical variable selection
                       # group number of the exposure variables is listed in the
                       # order that the variables appear as columns in Z matrix
                       groups = c(1,2,1,3)
                       # if I wanted to use fine-tuned parameters e.g:
                       # control.params = list(r.jump1 = 0.5, r.jump2 = 0.5)
                       # r.jump if varsel=FALSE
                       )


# The posterior inclusion probabilities (PIPs) for groups indicate that one of 
# the pollutants from Group 1 should be included. Looking at the exposures z1
# and z3 within Group 1, the results suggest that z1 is more likely to be associated
# with the outcome
bkmr::ExtractPIPs(fitkm_hier)


# We can explore the univariate relationship between each exposure and the outcome,
# where all of the other exposures are fixed to a particular percentile and all 
# covariates are held constant. The argument specifying the quantile at which to
# fix the other exposures is given by q.fixed. 
# Here we see that for z1 and z2, increasing values of the exposures is associated
# with increasing values of the outcome Y (for each of the other predictors in Z
# fixed to their 50th percentile). It also looks like z1 may have a nonlinear 
# relationship with the outcome

pred.resp.univar.hier <- bkmr::PredictorResponseUnivar(fit = fitkm_hier,
                                                        q.fixed = 0.5)

ggplot(pred.resp.univar.hier, 
       aes(z, est, ymin = est - 1.96*se, ymax = est + 1.96*se)
       ) + 
      geom_smooth(stat = "identity") + 
      facet_wrap(~ variable) +
      ylab("h(z)")


# We can explore the predictor-response function of a single predictor in Z for 
# the second predictor in Z fixed at various quantiles (and for the remaining 
# predictors fixed to a particular percentile). 
# The argument qs specifies a sequence of quantiles at which to fix the second 
# predictor.

pred.resp.bivar <- bkmr::PredictorResponseBivar(fit = fitkm_hier, min.plot.dist = 1)
pred.resp.bivar.levels <- bkmr::PredictorResponseBivarLevels(
                          pred.resp.df = pred.resp.bivar, 
                          Z = Z, 
                          qs = c(0.1, 0.5, 0.9))
# If the lines are parallel, there is no interaction between these two elements,
# if not, there is an interaction
ggplot(pred.resp.bivar.levels, aes(z1, est)) + 
    geom_smooth(aes(col = quantile), stat = "identity") + 
    facet_grid(variable2 ~ variable1) +
    ggtitle("h(expos1 | quantiles of expos2)") +
    xlab("expos1")


# We can compute the overall effect of the predictors, by comparing the value of 
# h (kernel function) when all of the predictors are at a particular percentile 
# and when all of them are at their 50th percentile. 
# The qs arg takes a sequence of values of quantiles and q.fixed the fixed quantile 
# (default is 50th percentile)

risks.overall <- bkmr::OverallRiskSummaries(fit = fitkm_hier, y = y, Z = Z, X = X, 
                                      qs = seq(0.25, 0.75, by = 0.05), 
                                      q.fixed = 0.5, method = "exact")
risks.overall
ggplot(risks.overall, aes(quantile, est, ymin = est - 1.96*sd, ymax = est + 1.96*sd)) + 
    geom_pointrange()


# Another summary of h that may be of interest would be to summarize the 
# contribution of an individual predictor to the response (single-predictor 
# health risks). For example, we may wish to compare risk when a single predictor
# in h is at the 75th percentile as compared to when that predictor is at its 
# 25th percentile, where we fix all of the remaining predictors to a particular 
# percentile. The two different quantiles at which to compare the risk are 
# specified using the qs.diff argument, and a sequence of values at which to
# fix the remaining pollutants can be specified using the q.fixed argument.
# The plot makes it easier to see.

risks.singvar <- bkmr::SingVarRiskSummaries(fit = fitkm_hier, y = y, Z = Z, X = X, 
                                      qs.diff = c(0.25, 0.75), #h(z_{q2}) - h(z_{q1})
                                      q.fixed = c(0.25, 0.50, 0.75),
                                      method = "exact")

# Here we see that a change in z2 from it's 25th to 75th percentile, where the 
# rest of the predictors are fixed at their 75th percentile (and the covariates x
# are constant), results in a ~1.16 estimated (additive) change in h(). 

# We also see that predictor z4 is not contributing to the "risk" and that
# z1 and z3 95% CI cross 0, meaning there is larger variation in their actual
# contribution to the outcome (it's not always/really contributing)
ggplot(risks.singvar, aes(variable, est, ymin = est - 1.96*sd, 
                          ymax = est + 1.96*sd, col = q.fixed)) + 
       geom_pointrange(position = position_dodge(width = 0.75)) + 
        coord_flip() +
        labs(x = "variable (from 25th to 75th percentile)")
  
 

# To make this notion a bit more formal, we may wish to compute specific ‘interaction’ 
# parameters. For example, we may want to compare the single-predictor health risks
# when all of the other predictors in Z are fixed to their 75th percentile to when
# all of the other predictors in Z are fixed to their 25th percentile. In the previous
# plot, this corresponds to subtracting the estimate represented by the red circle
# from the estimate represented by the blue circle. 

risks.int <- bkmr::SingVarIntSummaries(fit = fitkm_hier, y = y, Z = Z, X = X, 
                                 qs.diff = c(0.25, 0.75), # in z1
                                 qs.fixed = c(0.25, 0.75), # the rest of z
                                 method = "exact")
risks.int
 
# We see that not one single pollutant risk associated with a change in the exposure
# from its 25th to 75th percentile increases or decreases (all around 0 with
# a sd > |est|) when all other exposures are fixed at their 25th percentile, as 
# compared to when all other z are fixed at their 75th percentile, meaning no 
# "interaction" between them 
 


```
:::


## References
::: {.refs-small-text}
1) https://jenfb.github.io/bkmr/overview.html
:::