---
title: "Mediation analysis"
format: html
---

## What question does this answer?
How does an exposure affect an outcome via an intermediate variable?
Does the effect of exposure to outcome change now that we've introduced another
variable in the mix that the exposure does have an effect on?
Causal inference by finding an underlying mechanism under effect of the exposure,
and contributing to the outcome (be wary or saying something causes something,
experimental setup is critical there). A mediator explains the underlying mechanism 
of the relationship between the exposure and the outcome. 

## What is it doing?
We have an exposure that has an effect (at least theoretically a relationship)
with the outcome, and we have another variable that the exposure has an effect on
generally (remember to see assumptions below in @sec-when!) This other variable 
we'll call the mediator.
Effectively, we care about whether there is a difference in the exposure's
effect on the outcome now that we have introduced the mediator, or not.

## When can I use it? {#sec-when}

- There is a relationship between exposure and outcome (even if the regression
doesn't return significance, if there is a good theoretical reasoning, that's
enough). Without a relationship, there is nothing to mediate  
- There is a relationship between exposure and mediator, as is the idea that the
exposure affects the mediator that affects the outcome
- The outcome is binary -> **not a requirement**, just a package(s) and result 
interpretation convenience  
- Assumptions:  
  1) Must control for exposure-outcome confounding (i.e. include all covariates)  
  2) Must control for mediator-outcome confounding   
  3) Must control for exposure-mediator confounding   
  4) The exposure must not be affecting any of the mediator-outcome confounders 
(o/w C2 would be a mediator of E -> Y and is hence confounding the effect of M on Y)  


**Note #1:** In practice, some of the covariates may affect all of E, M and Y, and 
the covariates may also affect each other. None of this is problematic
and the covariate groups C1, C2, and C3 need not be distinguished from one another. What is
important is that the covariates included in the regression models above suffice to control for
exposure-outcome, mediator-outcome, and exposure-mediator confounding

**Note #2:** Another way to think about the fourth assumption is that there should 
be relatively little time between the exposure and the mediator. If a substantial
gap exists, then the fourth assumption requires that nothing on the pathway from
the exposure to the mediator itself also independently affects the outcome.
This assumption likely becomes increasingly less plausible the more time that 
elapses between the exposure and the mediator


## Reporting

- Total effect/TE is the effect of X on Y without M (Y~X)  
- Direct effect/NDE/ADE is the effect of X on Y with M (Y~X+M)  
- Indirect effect/NIE/ACME is total minus direct, and what you want to see whether it is 
significant, as that answers the question of whether X works through M to effect
Y, i.e. whether M affects Y, but X no longer affects Y as much, now that M is there. 

:::{.panel-tabset}

### Mediation

Since the sequential ignorability assumption is untestable, we recommend that 
the researchers conduct a sensitivity analysis via the `medsens` function, 
which can be applied to certain statistical models - **TBA**
In practice, you won't be 100% sure of whether you have included all confounders,
so sensitivity analyses to test for robustness of the in/direct estimates are necessary

- You can have the mediator model and the outcome model have different covariates
(as in, you can add more covariates to the outcome model)

```{r}
#| label: mediation_pckg

set.seed(2014)
data("framing", package = "mediation")

# The data captures subjects that were exposed to different media stories about 
# immigration and the authors investigated how their framing influences
# attitudes and political behavior regarding immigration policy. They propose anxiety
# as the mediating variable for the causal effect of framing on public opinion. 
# We first fit the mediator model where the measure of anxiety (emo) is modeled as
# a function of the framing treatment (treat - binary) and pre-treatment covariates (age, 
# educ, gender, and income). Next, we model the outcome variable, which is a binary
# variable indicating whether or not the participant agreed to send a letter about
# immigration policy to his or her member of Congress (cong_mesg - binary). The 
# explanatory variables of the outcome model include the mediator, treatment status,
# and the same set of pre-treatment variables as those used in the mediator model.
# In this example, the treatment is expected to increase the level of respondents'
# emotional response, which in turn is hypothesized to make subjects more likely 
# to send a letter to his or her member of Congress. 

med.fit <- lm(emo ~ treat + age + educ + gender + income, data = framing)
out.fit <- glm(cong_mesg ~ emo + treat + age + educ + gender + income,
                          data = framing, family = binomial("probit"))

# In general, as long as computing power is not an issue, analysts should
# estimate confidence intervals via the bootstrap (boot=TRUE) with more than 1000 
# resamples, which is the default number of simulations (sims)
# The bias-corrected and accelerated (BCa) intervals has better asymptotic properties 
# and is often recommended for the estimation of mediation effects (Preacher and Hayes 2008)
med.out <- mediation::mediate(med.fit, 
                   out.fit, 
                   treat = "treat", # treat is short for treatment aka exposure
                   mediator = "emo",
                   boot = TRUE,
                   boot.ci.type = "bca",
                   sims = 1000)

summary(med.out)
# ...
# Total Effect              0.095797    -0.030523     0.234075   0.136
# ...
# ACME (average)            0.083833     0.037913     0.144588  <2e-16 ***
# ADE (average)             0.011964    -0.097811     0.142607   0.794    
# Prop. Mediated (average)  0.875106     0.315506    37.469486   0.136    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Sample Size Used: 265


```
In this example, the estimated ACMEs are statistically significantly different 
from zero. The results suggest that the treatment in the framing experiment may 
have increased emotional response, which in turn made subjects more likely to send 
a message to his or her member of Congress. Here, since the outcome is binary, 
all estimated effects are expressed as the **increase in probability** that the 
subject sent a message to his or her Congress person.

The ACME is the estimated average increase in the dependent variable on average 
that arrives as a result of the mediators rather than 'directly' from the treatment.
The ACME (control) is the estimated average increase in the dependent variable
among the control group hat arrives as a result of the mediators rather than 
'directly' from the treatment.

If the outcome (and mediator) is continuous, ACME is in units of the outcome and
interpreting it is like for a standard linear regression (if data was scaled, 
then it is in standard deviations, not 1-unit changes).  
So if we had here 0.083 for ACME and the outcome was continuous, and we scaled 
at least the exposure and outcome, it would be interpreted as a 1-standard-deviation
increase in the exposure, then the pathway through the mediator "causes' the outcome 
to change by 0.083 standard deviations on average (rather than through 'just' the exposure).  
If it's just the exposure that's scaled, a 1-standard-deviation increase in the exposure, 
then the pathway through the mediator causes the outcome to change by 0.083 raw units.  
If it were only the outcome that was scaled, a 1-unit increase in the exposure would lead to the 
pathway through the mediator causing the outcome to change by 0.083 standard deviations.

This is the average effect for "everyone",  ignoring individual covariate differences.


### Medlfex

- The neWeight/neImpute model and the neModel model have to have the same 
covariates (based on the vignette)
- caveat: when wrapping medflex mediation in a function,
it searches for global environment variables, instead of inside the function's env


```{r}
#| label: medflex_pckg
 
set.seed(2014)
data("UPBdata", package = "medflex")

# The data is a subsample of 385 individuals from a survey study that involved the
# recruitment of individuals who divorced between March 2008 and March 2009 in
# four major courts in Flanders.
# It aimed to improve the quality of life in families during and after the divorce 
# by translating research findings into practical guidelines for separation 
# specialists (such as lawyers, judges, psychologists, welfare workers...) and by
# promoting evidence-based policy. The corresponding dataset (UPBdata) involves 
# those who responded to a battery of questionnaires related to romantic relationship 
# characteristics (such as adult attachment style) and breakup characteristics 
# (such as breakup initiator status, experiencing negative affectivity and 
# engaging in unwanted pursuit behaviors; UPB) (DeSmet, Loeys, and Buysse 2012). 
# Respondents were asked to imagine their former partner as well as possible and 
# to remember how they generally felt in their relationship before the breakup 
# when completing the attachment style questionnaire. The mediation hypothesis of
# interest concerned the question whether the level of emotional distress or 
# negative affectivity experienced during the breakup can be regarded as an 
# intermediate mechanism (M) through which attachment style towards the ex-partner
# before the breakup (X) exerts its influence on displaying UPBs after the breakup
# (Y) 

# As direct and indirect effects are most easily understood for a binary exposure, 
# we will use a dichotomized version of anxious attachment level (attbin) for 
# didactic purposes. Moreover, negative affectivity (negaff) has been standardized 
# (sd=1) to allow for easily interpretable effect estimates. The original att has 
# been scaled. The outcome variable unwanted pursuit behavior (UPB) indicates 
# whether (=1) or not (=0) the respondent has engaged in any unwanted pursuit 
#behaviors.

medFit <- glm(negaff ~ factor(attbin) + gender + educ + age,
              family = gaussian, 
              data = UPBdata)

expData <- medflex::neWeight(medFit)

# The new variables attbin0 and attbin1 correspond to hypothetical exposure 
# values x and x∗, respectively. By convention, the index ‘0’ is used for 
# parameters (and corresponding auxiliary variables) indexing natural direct 
# effects, whereas the index ‘1’ is used for parameters indexing natural indirect 
# effects in the natural effect model. To shorten code, one can instead choose to 
# directly specify the formula, family and data arguments in neWeight.

expData <- medflex::neWeight(negaff ~ factor(attbin) + gender + educ + age, 
                             family = gaussian, 
                             data = UPBdata)

# As an alternative, robust standard errors based on the sandwich estimator 
# (Liang and Zeger 1986) can be requested by setting se = "robust". Calculation 
# of these standard errors is less computer-intensive and is available for natural 
# effect models with working models fitted via the glm function

neMod1 <- medflex::neModel(UPB ~ attbin0 + attbin1 + gender + educ + age, 
                           family = binomial("logit"), 
                           expData = expData,
                             se = "robust")


# The summary table of the resulting natural effect model object provides these 
# bootstrap standard errors along with corresponding Wald-type z statistics and
# p values
summary(neMod1)
# you could also see the total effect by using
medflex::neEffdecomp(neMod1)

```

A counterfactual model reporting:  
Exponentiating the model parameter estimates provides estimates that can be 
interpreted as odds ratios. For instance, for a subject with baseline covariate 
levels C, altering the level of anxious attachment from low (=0) to high (=1), 
while controlling negative affectivity at levels as naturally observed at any 
given level of anxious attachment x, increases the odds of displaying unwanted 
pursuit behaviors with a factor $e^{\beta_1} = e^{0.3959} = 1.49$ (natural direct effect)

Altering levels of negative affectivity as observed at low anxious attachment scores
to levels that would have been observed at high anxious attachment scores, while
controlling their anxious attachment score at any given level x, increases the 
odds of displaying unwanted pursuit behaviors with a factor  $e^{\beta_2} = e^{0.352} = 1.42$ 
(natural indirect effect).

```{r}
#| label: medflex_pckg_conf_int

# Wald-type confidence intervals can be obtained by applying the confint function
# to the natural effect model object. The confidence level defaults to 95%, but 
# can be changed via the level argument. By exponentiating the intervals on the 
# logit scale, we can obtain the corresponding 95% confidence intervals (based on 
# the robust standard errors) on the odds ratio scale: If standard errors are 
# obtained via the bootstrap procedure, bootstrap confidence intervals are returned.
# The default type is calculated based on a first order normal approximation
# (type = "norm"), but other types of bootstrap confidence intervals (such as basic bootstrap,
# bootstrap percentile and bias-corrected and accelerated confidence intervals) can be obtained
# by setting the type argument to the desired type.

exp(confint(medflex::neEffdecomp(neMod1)))
```


```{r}
#| label: medflex_cont_exp

set.seed(2014)
# we can have continuous exposures, where instead of enumerating all
# exposure levels to construct auxiliary variables x and x∗
# for each subject, Vansteelandt et al. (2012b) proposed to draw specific quantiles
# from the conditional density of the exposure given baseline covariates. 
# By default, these hypothetical exposure levels are drawn from a linear
# model for the exposure, conditional on a linear combination of all covariates 
# specified in the working model. 
# If one wishes to use another model for the exposure, this default model 
# specification can be overruled by referring to a fitted model object in the xFit
# argument. Misspecification of this sampling model does not induce bias in the 
# estimated coefficients and standard errors of the natural effect model

# Both neWeight and neImpute allow to choose the number of draws to sample from this
# conditional density via the nRep argument (which defaults to 5), simulations 
# have shown the gain to be ignorable when choosing more than 5 draws 

expData_cont_exp <- medflex::neWeight(negaff ~ att + gender + educ + age,
                    family = gaussian, 
                    data = UPBdata, 
                    nRep = 5)

neMod_cont_exp <- medflex::neModel(UPB ~ att0 + att1 + gender + educ + age,
                          family = binomial("logit"), 
                          expData = expData_cont_exp, 
                          se = "robust")
summary(neMod_cont_exp)


```

For a subject with baseline covariate levels C, the direct and indirect effects of
one standard deviation increase in anxious attachment level correspond
to an increase in the odds of displaying unwanted pursuit behaviors by a factor
$e^{\beta_1} = e^{0.29578} = 1.34$ (34% increase in odds) (NDE) and $e^{\beta_2} = e^{0.18324} = 1.2$ (NIE),
regardless of the initial level

If the outcome was continuous, it would be interpreted the same as it was in the
mediation package, only with the caveat that the output medflex gives is conditional
on baseline covariate levels C. The difference between conditional and marginal
is negligible if there is no interaction between the variables. medflex handles
interactions in models TBA.

Usecases for between neWeight and neImpute TBA (does the same thing but different
approaches; weighted-based approach amounts to fitting a working model for the
mediator, whereas the imputation-based approach requires fitting a mean model
for the outcome)
:::

You can report proportion mediated, but without the confidence intervals and p-value
(beware, it can be >1 or <0 due to directionality, which does not make sense)





## References
:::{.refs-small-text}
1) https://doi.org/10.1146/annurev-publhealth-032315-021402 - Mediation Analysis: A Practitioner's Guide
2) https://library.virginia.edu/data/articles/introduction-to-mediation-analysis - traditional overview of mediation
3) https://www.sciencedirect.com/science/article/pii/S0013935126009382?via%3Dihub#sec2 - paper uses mediation and IRR to report
4) https://spout.ussg.indiana.edu/CRAN/web/packages/mediation/vignettes/mediation.pdf
5) https://cran.r-project.org/web/packages/medflex/vignettes/medflex.pdf
6) https://stats.stackexchange.com/questions/104692/comprehending-output-from-mediation-analysis-in-r
7) https://pmc.ncbi.nlm.nih.gov/articles/PMC10615709/ - example where E,M,Y are binary
8) https://arxiv.org/pdf/1904.08515 - clarifying casual for the researcher
9)https://pmc.ncbi.nlm.nih.gov/articles/PMC10415648/ - comparison of 2 packages
:::